Re-run, Repeat, Reproduce, Re-use, Replicate:\\Transforming Code into Scientific Contributions

### 2017-06-13 main.tex
Reworked Reproducible (R^3).

% !TeX program = pdflatex
\documentclass[a4paper,11pt]{article}

@@

%Replication is a cornerstone of science. 
Reproducibility is a cornerstone of science. 
% Since science aims to discover general principles of how the world functions, 
% * <nicolas.rougier@inria.fr> 2017-04-24T15:49:43.902Z:
% 
% > Since science aims to discover general principles of how the world functions, 
% 
% 
% C'est un peu une déclaration à l'emporte pièce et je soupçconne qu'il y a une littérature énorme sur la définition de la science. Et à moins de vouloir s'aliéner toute l'astophysioque et la cosmologie, on pourrait au minimum étendre à l'univers.
% 
% ^ <fabien.benureau+overleaf@gmail.com> 2017-04-25T13:29:01.309Z:
% 
% Reopened by editing the source!
% From [wikipedia](https://en.wikipedia.org/wiki/World): "In a philosophical context, the world is the whole of the physical Universe, or an ontological world."
% 
% ^ <nicolas.rougier@inria.fr> 2017-04-25T13:45:04.565Z:
%
% Mouais.
%
% ^.
If an experimental result cannot be obtained more than once,
% it might be because of a fluke, a bias, a bug, a false positive or even a hack.
%it becomes, at best, nothing more than food for thought for future research. 
it merely becomes, at best, an observation that
may feed future research \cite{Mesirov:2010,osc:2015}. 
Replication issues have received increased attention in recent years,
in the domain of medicine and psychology in particular \cite{Iqbal:2016}.
One could think that computer science would mostly be shielded from such issues.
But precisely because it is easy to believe that
if a program runs once and gives expected results it will do so forever, 
crucial steps to transform working code into meaningful scientific contributions are rarely undertaken \citep{Sandve:2013,Schwab:2000}. 
Computer science is plagued by replication problems,
in part, precisely because it seems impervious to them.\\

In fact, a program can fail as a scientific contribution
in many different ways for many different reasons.
Borrowing the terms coined by \citeauthor{Goble:2016} \citep{Goble:2016},
for your program to contribute to Science,
you have to make sure it is
re-runnable (R$^1$),
repeatable (R$^2$),
reproducible (R$^3$),
re-usable (R$^4$)
and replicable (R$^5$).\\

@@

To be clear, it is impossible to write future-proof code,
as the best efforts can be stymied by the smallest change in one of the dependencies.
At the same time, modernizing an unmaintained ten-year-old code
can reveal itself to be an arduous and expensive undertaking,
with each change risking to affect the semantics of the program.

Rather trying to predict the future or painstakingly dusting off old code, 
an often easier solution is to recreate the old execution environment.
For this to happen however, 
the dependencies in terms of systems, software and libraries must be made clear enough.

A \emph{re-runnable} code is one that describes an execution environment in which it is executable with enough details to be recreated.
As shown by \citep{Collberg:2016}, this is far from being neither obvious nor easy.

@@

In our case, the R$^0$ version of our tiny walker seems to imply
that any version of Python would be fine.
This not the case because it uses the print {\em instruction} and the {\tt xrange} operator, both specific to Python 2.
The print {\em instruction}, available in Python 2 (still widely used, will stop to be supported in 2020), 
has been deprecated in Python 3 (first released in 2008, almost a decade ago)
in favor or a  print {\em function}.
The {\tt xrange} operator has been replaced by the {\tt range} operator in Python 3.
In order to try to future-proof the code a bit, we might as well target Python 3, as is done in the R$^1$ version.
Incidentally, it remains compatible with Python 2.\\

But whatever the version chosen,
the crucial step here is to document it.

@@

% * <fabien.benureau+overleaf@gmail.com> 2017-04-27T06:18:15.753Z:
% 
% - 
% 
% Ok, so I am not sure that the parameter talk should be here. It should be in reproducible. 
% 
% ^.

The code is running and producing the expected results. 
The next step is to make sure that you can produce the same output over successive runs of your program, should you want to. 
In other words, the next step is to make your program deterministic,  producing {\em repeatable} output. 
Such repeatability is useful. 
If a run of the program produces a particularly puzzling results,  it allows you to scrutinize any step of the execution of the program by re-running it again. 
It is also the easiest way to prove that the program did indeed produce the published results.
Such repeatability is not always possible or easy \citep{Diethelm:2012, Courtes:2015}.
But for sequential programs not depending on exotic hardware, it often comes down to controlling the initialization of the pseudo-random number generators (RNG).\\

For our program, that means setting the seed of the {\tt random} module. We may also want to save the output of the program to a file, so that we can easily verify that consecutive runs do produce the same output; eyeballing differences is unreliable and  time-consuming (and therefore won't be done systematically).

@@

Setting seeds is not without danger. 
Using 439 as a seed in the previous program would result in ten consecutive +1 steps, which---although a perfectly valid random walk---lend itself to gross misinterpretations of the overall dynamic of the algorithm. 
Verifying that the qualitative aspects of the results and the conclusion we make are not tied to a specific initialization of the pseudo-random generator is an integral part of any scientific undertaking in computational science; 
this is usually done by repeating the simulations multiple times with different seeds.

@@

% \clearpage
\section*{Reproducible (R$^{\mathbf 3}$)}

The R$^2$ code seems fine enough, but it hides several problems that come to light when trying to {\em reproduce} results.
In the strongest sense, a code is said to be {\em reproducible} if the same results it produced in the past can be re-obtained by someone else.
As explained by \citeauthor{Mesnard:2016} \citep{Mesnard:2016}, reproducibility is harder than you think. 
For instance, our R$^2$ program will not produce the same results for everyone. It will, because it is repeatable, produce the same results over repeated executions.
But it will not do so over different execution environments. The cause is found in a change that occurred in the pseudo-random number generator between CPython 3.2 and CPython 3.3. 
Executed with CPython 3.2 or older, the code will produce the sequence 1, 2, 1, 2, 3, 4, 5, 6, 7, 6. But with CPython 3.3 or newer, it will produce 1, 2, 1, 0, 1, 0, 1, 0, -1, 0.
Reproducibility impose to specify dependencies more precisely than re-runnability: the exact versions that were used must be documented. The platform should also be described as well: CPU architecture, operating system version, endianness, etc., as well as any defaults you may have set that might have an impact.
Because of the increasing complexity of our computational stacks, retrieving and deciding what is pertinent might be difficult. A best effort must be made.\\

Here, in addition to specifying the exact Python version and platform details, a short unit test is provided, that allows the code to verify its own reproducibility. Should this test fail, then there is little hope of reproducing the results. If it does not fail, problems may still be present. 
A second problem present in the R$^2$ program is that it does not store the parameters used along with the results. Therefore, should the code change after the production of the results, someone provided with the result file will not be able to know which seed was used to produce it. In R$^3$, the seed is stored in the name of the result file.

@@

\begin{code}{\textbf{\textsc{Listing 4:}} Random walk (R$^3$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C.Y. Benureau
# Release under the BSD 2-clause license
# Tested with CPython 3.6 / macOS 10.12.4 / 64 bits architecture
import random
from itertools import accumulate

@@

Finally, a crucial condition to making a code reproducible is making it available. As shown in \citep{Collberg:2016}, code is often not available, or available upon request. The latter may seem sufficient, but changes in email address,  career or retirement can make the code just as unavailable. Code \emph{and input and result files} should be available with the published article, as supplementary data, or through a DOI link to a scientific repository such as Figshare or Zenodo.
Providing results files is important: it allows other researchers to easily compare the results they obtains with the one used to generate the figure in the article. Comparing figures directly can be cumbersome and often only allows to conclude about a qualitative similarity.\\

To recap, reproducibility imposes stronger conditions than re-runnability. Dependencies and platforms must be described as precisely and a specifically as possible. Unit tests are a good way to embed self-diagnostics of reproducibility in the code. Results must be accompanied with the parameters and inputs used to produce them. And the code, and the data files behind the figure must be made available.

% \clearpage
\section*{Re-usable (R$^{\mathbf 4}$)}

Making your program re-usable means it can be used by people outside your lab,
using possibly different environment, different parameters or different data.
If people start using your program, they will most likely report bugs or malfunctions they encountered while using your program.
If you're lucky enough, they will even propose you bugfix, hence improving the overall quality and correctness of your software.
In other words, if your program is re-usable and is re-used, it is alive.
As such, and as for any other living being, it will evolve and mutate to adapt to its environment.
This process will ensure long-term reproducibility to the extent people continue using and maintaining the program.\\

@@

>>>>

% !TeX program = pdflatex
\documentclass[a4paper,11pt]{article}
% \usepackage[utf8]{inputenc}

@@

Replicability is a cornerstone of science. 
If an experimental result cannot be re-obtained by an independent party,
% it might be because of a fluke, a bias, a bug, a false positive or even a hack.
%it becomes, at best, nothing more than food for thought for future research. 
it merely becomes, at best, an observation that
may feed future research (\cite{Mesirov:2010,osc:2015}). 
Replication issues have received increased attention in recent years,
in the domain of medicine and psychology in particular (\cite{Iqbal:2016}).

One could think that computer science would mostly be shielded from such issues.
But precisely because it is easy to believe that
if a program runs once and gives expected results it will do so forever, 
crucial steps to transform working code into meaningful scientific contributions are rarely undertaken \citep{Sandve:2013,Schwab:2000}. 
Computer science is plagued by replication problems,
in part, precisely because it seems impervious to them.

In fact, a program can fail as a scientific contribution
in many different ways for many different reasons.
Borrowing the terms coined by \citeauthor{Goble:2016} \citep{Goble:2016},
for your program to contribute to Science,
you have to make sure it is
re-runnable (R$^1$),
repeatable (R$^2$),
reproducible (R$^3$),
re-usable (R$^4$)
and replicable (R$^5$).

@@

To be clear, it is impossible to write future-proof code,
and the best efforts can be stymied by the smallest change in one of the dependencies.
At the same time, modernizing an unmaintained ten-year-old code
can reveal itself to be an arduous and expensive undertaking,
with each change risking to affect the semantics of the program.

Rather trying to predict the future or painstakingly dusting off old code, 
an often easier and less dangerous solution is to recreate the old execution environment.
For this to happen however, 
the dependencies in terms of systems, software and libraries must be made clear enough.

A \emph{re-runnable} code is one that describes an execution environment in which it is executable with enough details to be recreated.
As shown by \citep{Collberg:2016}, this is far from being either obvious or easy.

@@

In our case, the R$^0$ version of our tiny walker seems to imply
that any version of Python would be fine.
This not the case because it uses the print {\em instruction} and the {\tt xrange} operator, both specific to Python 2.
The print {\em instruction}, available in Python 2 (a version still widely used; support is scheduled to stop in 2020), 
has been deprecated in Python 3 (first released in 2008, almost a decade ago)
in favor or a  print {\em function}.
The {\tt xrange} operator has been replaced by the {\tt range} operator in Python 3.
In order to try to future-proof the code a bit, we might as well target Python 3, as is done in the R$^1$ version.
Incidentally, it remains compatible with Python 2.

But whatever version is chosen,
the crucial step here is to document it.

@@

The code is running and producing the expected results. 
The next step is to make sure that you can produce the same output over successive runs of your program. 
In other words, the next step is to make your program deterministic,  producing {\em repeatable} output. 
Such repeatability is useful. 
If a run of the program produces a particularly puzzling results,  it allows you to scrutinize any step of the execution of the program by re-running it again with extraneous prints, or inside a debugger. 
It is also the easiest way to prove that the program did indeed produce the published results.
Such repeatability is not always possible or easy \citep{Diethelm:2012, Courtes:2015}.
But for sequential programs not depending on exotic hardware, it often comes down to controlling the initialization of the pseudo-random number generators (RNG).\\

For our program, that means setting the seed of the {\tt random} module. We may also want to save the output of the program to a file, so that we can easily verify that consecutive runs do produce the same output: eyeballing differences is unreliable and  time-consuming---and therefore won't be done systematically.

@@

Setting seeds is not without danger. 
Using 439 as a seed in the previous program would result in ten consecutive +1 steps, which---although a perfectly valid random walk---lend itself to a gross misinterpretation of the overall dynamic of the algorithm. 
Verifying that the qualitative aspects of the results and the conclusion we make are not tied to a specific initialization of the pseudo-random generator is an integral part of any scientific undertaking in computational science; 
this is usually done by repeating the simulations multiple times with different seeds.

@@

%\clearpage
\section*{Reproducible (R$^{\mathbf 3}$)}

% Reproducible definition + R^2 not reproducible
The R$^2$ code seems fine enough, but it hides several problems that come to light when trying to {\em reproduce} results.
A result is said to be \emph{reproducible} if another researcher can take the original code and input data, execute it, and re-obtain the result (\cite{Peng:2006}). 
As explained by \citeauthor{Mesnard:2016} \citep{Mesnard:2016}, reproducibility is hard. 
For instance, the R$^2$ program will not produce the same results all the time. It will, because it is repeatable, produce the same results over repeated executions.
But it will not do so over different execution environments. The cause is found in a change that occurred in the pseudo-random number generator between Python 3.2 and Python 3.3. 
Executed with Python 2.7 to 3.2, the code will produce the sequence 1, 2, 1, 2, 3, 4, 5, 6, 7, 6. But with Python 3.3 to 3.6, it will produce 1, 2, 1, 0, 1, 0, 1, 0, -1, 0.\\

% Reproducible definition + R^2 not reproducible
Because any dependency of your program---to the most basic one, the language itself---can change its behavior from one version to another, executability (R$^1$) and determinism (R$^2$) are necessary but not sufficient for reproducibility. The exact execution environment used to produce the results must also be specified---rather than the broadest set of environments where the code can be executed. In other words, assertions such as "the results were obtained with Python 3.6.1" are more valuable, in a scientific context, than "the program should work with Python 3.3.x and above".\\

% Reproducibility is hard
Because of the increasing complexity of our computational stacks, retrieving and deciding what is pertinent (CPU architecture? Operating system version? Endianness?) might be difficult.
See Appendix \ref{appendix:executionenvs} for a discussion.\\

% Reproducibility in practice.
Having precisely described the execution environment, reproducibility is technically obtained. However, \emph{reproducibility in practice} requires more than that.

% Saving Parameters
The R$^2$ program uses a random seed but does not keep a trace of it except in the code. Therefore, should the code change after the production of the results, someone provided with the result file will not be able to know which seed was used, and would need to iterate through all possible random seeds, which is not possible in practice. R$^3$ saves the seed in the name of the result file. More generally, results should always come alongside their context, i.e. an exhaustive list of the parameters used and a description of the execution environment.\\

% Data behind graphs
Graphs in articles are powerful way to communicate complex information. They are imprecise and cumbersome, and sometimes plain unusable, however when it comes to checking that the reproduced data match the one published in the article. A good figure should always be accompanied by its underlying data to allow straightforward comparisons.\\

% Unit test for R^3
Another good practice is to make the code self-verifiable. In R$^3$, a short unit test is provided, that allows the code to verify its own reproducibility. Should this test fail, then there is little hope of reproducing the results. Of course, passing the test is no definitive proof.\\

% Reproducibility implies availability 
It should be obvious by now but \emph{reproducibility implies availability} As shown in \citep{Collberg:2016}, code is often unavailable, or only available upon request. While the latter may seem sufficient, changes in email address, changes in career, retirement, a busy inbox or poor archiving practices can make a code just as unreachable. Code \emph{and input and result files} should be available with the published article, as supplementary data, or through a DOI link to a scientific repository such as Figshare or Zenodo\footnote{Online code repository such as GitHub \emph{are not} scientific repositories, and may disappear, change name, etc. at any moment. Direct links to them are not perpetual and should be avoided.}.\\

@@

\begin{code}{\textbf{\textsc{Listing 4:}} Random walk (R$^3$)}
# Copyright (c) 2017 Nicolas P. Rougier and Fabien C. Y. Benureau
# Release under the BSD 2-clause license
# Tested with CPython 3.6 / macOS 10.12.4 / 64 bits architecture
import random
from itertools import accumulate

@@

% Summary
To recap, reproducibility implies re-runnability or repeatability and availability, yet  imposes additional conditions. Dependencies and platforms must be described as precisely and a specifically as possible. Parameters values and inputs should be an integral part of the result files. The data behind graphs must be published. Unit tests are a good way to embed self-diagnostics of reproducibility in the code. Reproducibility is hard, yet tremendously necessary.

\clearpage
\section*{Re-usable (R$^{\mathbf 4}$)}

Making your program re-usable means it can be used by people outside your lab,
using possibly different environment, different parameters or different data.
If people start using your program, they will most likely report bugs or malfunctions they encountered while using your program.
If you're lucky enough, they will even propose you bugfixes, hence improving the overall quality and correctness of your software.
In other words, if your program is re-usable and is re-used, it is alive.
As such, and as for any other living being, it will evolve and mutate to adapt to its environment.
This process will ensure long-term reproducibility to the extent people continue using and maintaining the program.\\

@@

\newpage 
\appendix

\section{Appendix: Recording Execution Environments}
\label{appendix:executionenvs}
Talk about provenance-tracking and its problems (endianess, "as well as any defaults you may have set that might have an impact.")

Talk about saving VMs, Docker images and its problems.
